---
title: "Banana Quality Prediction"
output:
  html_document:
    theme: null
    highlight: textmate
    df_print: paged
    css: ../assets/styles.css
---

This project develops machine learning models to classify banana quality in supermarket supply chains. The goal is to reduce operational losses by identifying low-quality bananas early using features such as sweetness, acidity, and ripeness. The code covers data preparation, model training, and model evaluation.

# **Step 1: Load and Clean the Data**

Load dataset, standardize target column (quality), and convert to factor for classification
```{r setup, include=FALSE}
library(tidyverse)
library(caret)
library(rpart)
library(class)
library(rpart.plot)
library(ggplot2)
```

```{r load_data, echo=TRUE}
banana <- read.csv("../data/banana_quality.csv")

names(banana)[names(banana) == "Quality"] <- "quality"
banana$quality <- as.factor(banana$quality)
```

Convert quality into a factor (classification labels: good or bad)
```{r convert_quality_to_factor, echo=TRUE}
banana$quality <- as.factor(banana$quality)
```

# **Step 2: Check the Data**

Inspect data structure, types, and missing values before modeling


Look at the first few rows of the dataset
```{r head, echo=TRUE}
head(banana)
```

Check the structure (shows column types)
```{r structure, echo=TRUE}
str(banana)
```

Check for missing values in each column
```{r missing_data, echo=TRUE}
colSums(is.na(banana))
```

Summary statistics (min, max, mean, etc.)
```{r summary_stats, echo=TRUE}
summary(banana)
```


# **Step 3: Split the Data**

Split data into training (80%) and test (20%) sets to assess performance on unseen observations.

```{r set_seed, echo=TRUE}
# Set seed for reproducibility (so we always get the same split)
set.seed(123)
```

```{r spliy_data, echo=TRUE}
# Split the data: 80% for training, 20% for testing
train_index <- createDataPartition(banana$quality, p = 0.8, list = FALSE)
train_data <- banana[train_index, ]
test_data <- banana[-train_index, ]
```



# **Train and Test Data**

Train and evaluate **Logistic Regression, KNN, and Decision Tree models** on the same data for comparison.

# **Model 1: Logistic Regression**
Train LR Model

```{r train_lr_model, echo=TRUE}
# Train the logistic regression model
log_model <- glm(quality ~ ., data = train_data, family = "binomial")
```

```{r lr_summary, echo=TRUE}
# View model summary to see feature importance
summary(log_model)
```

All features are statistically significant. Positive coefficients increase the likelihood of “Good” quality, while acidity is negatively associated. The model shows strong fit.

**Evaluate model performance on the test set using a confusion matrix.**

```{r lr_confusion_matric, echo=TRUE}
# Make predictions on the test data
log_probs <- predict(log_model, test_data, type = "response")

# Convert probabilities to class labels using 0.5 cutoff
log_preds <- factor(ifelse(log_probs > 0.5, "Good", "Bad"), levels = c("Bad", "Good"))

# Evaluate model performance with a confusion matrix
confusionMatrix(log_preds, test_data$quality)
```

The model achieves 87.2% accuracy, with balanced sensitivity and specificity, indicating no strong bias toward either class.

# **Model 2: K-Nearest Neighbors (KNN)**

Scale features and apply KNN to classify banana quality based on nearest neighbors.

```{r train_knn, echo=TRUE}
# Remove the 'quality' column before scaling the numeric features
train_x <- train_data[, -which(names(train_data) == "quality")]
test_x <- test_data[, -which(names(test_data) == "quality")]

# Scale the training data
train_x_scaled <- scale(train_x)

# Scale the test data using the same parameters as the training set
test_x_scaled <- scale(
  test_x,
  center = attr(train_x_scaled, "scaled:center"),
  scale = attr(train_x_scaled, "scaled:scale")
)

# Extract the labels (target variable)
train_y <- train_data$quality
test_y <- test_data$quality

# Run KNN with k = 5
knn_preds <- knn(train = train_x_scaled, test = test_x_scaled, cl = train_y, k = 5)

```

```{r knn_confusion_matrix, echo=TRUE}
# Evaluate performance using confusion matrix
confusionMatrix(knn_preds, test_y)
```

```{r knn_heatmap, echo=TRUE}
cm <- confusionMatrix(knn_preds, test_y)

cm_df <- as.data.frame(cm$table)

ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), color = "white", size = 6) +
  scale_fill_gradient(low = "#FFCB05", high = "#00274C") +
  theme_minimal() +
  labs(title = "KNN Confusion Matrix")
```
The KNN model achieves 97.9% accuracy, significantly outperforming logistic regression with consistently high sensitivity and specificity.

# **Model 3: Decision Tree**

Train a decision tree to classify banana quality and evaluate an interpretable, rule-based model.

```{r train_dt, echo=TRUE}
# Train the decision tree model

tree_model <- rpart(quality ~ ., data = train_data, method = "class")

# Make predictions on the test set
tree_preds <- predict(tree_model, test_data, type = "class")

```

```{r plot_tree_basic, echo=TRUE}
rpart.plot(
  tree_model,
  type = 2,
  extra = 104,
  fallen.leaves = TRUE,
  tweak = 1.2
)

```
The tree uses feature-based splits such as sweetness, ripeness, and harvest time to classify quality through clear, interpretable rules.

```{r dt_confusion_matrix, echo=TRUE}
# Evaluate the decision tree performance
confusionMatrix(tree_preds, test_data$quality)

```
The decision tree achieves 88.6% accuracy, slightly outperforming logistic regression but below KNN. It shows balanced sensitivity (89.7%) and specificity (87.4%) and offers interpretable decision rules.


### Interpretation:
The decision tree model performs well in identifying both good and bad bananas and provides the added benefit of being **easy to interpret**. The model’s decision rules can be visualized and explained to non-technical stakeholders. 

While it does not outperform KNN in accuracy, it may be preferred in situations where **model transparency** and **explainability** are important.


# **Model Summary: Key Takeaways**

All models perform well with distinct strengths:

- **Logistic Regression**: Simple, fast baseline with lower accuracy.
- **KNN**: Highest accuracy and best for minimizing prediction errors.
- **Decision Tree**: Balanced performance with strong interpretability.

Overall, KNN is best for accuracy, the decision tree is best for interpretability, and logistic regression is best for speed and simplicity.







